# Working through a dataset 

```{r, echo = F}
knitr::opts_chunk$set(comment = "", prompt = T, class.source="Rchunk", class.output="Rout")
```

## Solving an outbreak! 

We first need to load in the [**packages**]{style="color:purple"} that we will be using and our [**data**]{style="color:purple"}.

::: {#teaching_moment_3 .yellowbox}

ðŸ” **What are packages?**


One of the great things about R is that there are lots of pre-installed commands that we can access or ones that others help to create and we can access. These are our **packages** that we access with the `library()` command.

Packages can help us inspect our data using simpler language. We are using the packages **tidyverse** @tidyverse2019 and **ggmap** @R-ggmap.

:::

```{r, message=F, warning=F}
library(tidyverse)
library(ggmap)
source(here::here("setup/setup.R"))
```


::: {#teaching_moment_3 .yellowbox}

ðŸ’¡ If `library(tidyverse)` or `library(ggmap)` is not working, then the packages may not be installed. To do this, simply run `install.packages(c("tidyverse", "ggmap"))` in the **console** or in the **code chunk**. 

:::


## What does `outbreak_data` look like ?

The number one step in data analysis is looking at the data we have to work with. For this, we are going to use our first pre-installed command called `colnames()`.

```{r}
colnames(outbreak_data)
```

- Can you find `outbreak_data` in the environment pane?
- What happens if you click on it there?

## Different kinds of data

In `outbreak_data` we have several kinds of information and it's important to identify what these are !

-   How would you describe what data we have in `Firstname`?

-   How is this different to `Weight` ?

-   What are some terms we could use to define these ?

To look specifically at **one column** we can **select** the column using a **pipe** or `%>%`. 

```{r}
outbreak_data %>% select(height)
```

## Visualising data

The types of data we have to work with will determine how it is best to visualise !

A [scatter plot]{style="color:purple"}: 

```{r}
mpg %>% 
  ggplot(aes(x = displ, y = cty)) + 
  geom_point(size = 3)+
  labs(y = "", x = "")
```

Notice both the x and the y-axis are things we can count or measure.

A [boxplot]{style="color:purple"}: 

```{r}
mpg %>% 
  ggplot(aes(x = class, y = cty)) + 
  geom_boxplot()+
  labs(y = "", x = "")
```

Notice the y-axis are things we count or measure but the x-axis are specific groups.

## Plotting our data

We will be using a function called `ggplot()`. This is a fantastic tool we can use to make amazing graphs !

Like any graph we need to decide:

1.  What do we want on the x-axis?
2.  What do we want on the y-axis?
3.  What kind of graph do we want to make?

Other ideas like colour, design, groups ... we'll get to this later! 

```{r, fig.align='center'}
ggplot(data = outbreak_data,
         aes(x = class, y = height))
```

These are both pieces of data represented by numbers - how else might we record data?

### How to ask ggplot to plot `outbreak_data`

```{r, fig.align='center'}
outbreak_data %>% 
  ggplot(aes(x = age, y = weight)) +
  geom_point()
```

Next we choose what kind of graph we need

### We can use `geom_point()` or `geom_boxplot()`

Let's first try `geom_point()`. We use the `+` sign to add it to our first ggplot command. 

```{r, fig.align='center'}
outbreak_data %>% 
  ggplot(aes(x = class, y = height)) +
  geom_point()
```

Where would we put `geom_boxplot()` ?

### Now swap <u>height</u> for <u>temperature</u> on the y-axis. 

```{r fig.width=7, fig.height=4, fig.align='center'}
ggplot(data = outbreak_data,
         aes(x = class, y = temperature))+
  geom_point()
```

## Let's customise! 

Can you change size, shape and colour...what kinds of plots can you make ?

```{r}
ggplot(data = outbreak_data,
       aes(x = class, y = temperature))+
  geom_point(size = 4, shape = 5, colour = "blue")
  
```

### Boxplots to create summaries 

Can you make a **boxplot** with **Class** on the x-axis and **Temperature** on the y-axis? A boxplot summarising <u>temperature</u> in the four <u>classes</u>. 

```{r fig.width=7, fig.height=4, fig.align='center'}
ggplot(data = outbreak_data,
       aes(x = class, y = temperature))+
  geom_boxplot()
```


::: {#teaching_moment_4 .yellowbox}

ðŸ’¡ Boxplots provide us with an informative summary of the data in our groups
-   The center line gives the median
-   the box below is 25% of the data below the median and the box below is 25% of the data above
    
:::

\

::: {#teaching_moment_5 .yellowbox}

ðŸ’¡ If we don't know what R is doing, we can always get [help !]{style="color:purple"}
    
:::

\

```{r, message=TRUE}
?geom_boxplot()
```


### How can we use fill to learn more ?

The `fill` can be specified as [aesthetics]{style="color:purple"} like the x and y axis.
Can you specify the `fill` with our `Sickness` information about each student ?

```{r}
ggplot(data = outbreak_data,
       aes(x = class, y = temperature, fill = sickness)) +
  geom_boxplot()
```

### What does temperature tell us about how the students were feeling? 

-   Were the temperatures the same for students who were sick and those who weren't

-   Where the sick students in all classes

-   Why do people get a temperature when they get sick?

### Time to give our plot a title

```{r fig.align='center'}
ggplot(data = outbreak_data,
         aes(x = class, y = temperature, fill = sickness))+
  geom_boxplot()+
  labs(title = "Make a title here !")
```

Make it something informative that describes the plot you've just made and why people would be interested to look at it

## Generating summaries

This symbol `%>%` is used to send our data into another command such as `summarise()`. 

As the name would suggest, we can generate summaries, such as calculating the [median]{style="color:purple"}. 

```{r}
outbreak_data %>% 
  summarise(median = median(temperature))
```


::: {#teaching_moment_5 .yellowbox}

ðŸ’¡ We can also calculate other summary statistics in R. Such as: 

outbreak_data %>% 
  summarise(median_age = median(age), 
            min_age = min(age),
            max_age = max(age),
            mean_age = mean(age))
    
:::

\


## We can also use `group_by()` with multiple groups

Just like our boxplot, we want to know the median temperature of the students, grouped by their `Class` and `Sickness` status. Try to do this by adding `group_by(Class, Sickness)`. 

```{r}
outbreak_data %>% 
  group_by(class, sickness) %>% 
  summarise(median = median(temperature))
```

How can we relate this to the boxplot we made earlier ?

